Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
More edits

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}(#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made one version, the data in the other must be updated so the
two continue to represent the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, 
programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses can be
written in one of several domain-specific bidirectional languages that 
guarantee various \emph{lens laws} constraining how the \kw{put} and
\kw{get} functions must fit together. \bcp{See my email about deemphasizing
  ``storing'' multiple formats.}

Recent work~\cite{optician}\saz{removed ``we'' here to maintain the double-blind
  review standard.} developed a tool called Optician to synthesize the
class of {\em bijective lenses} that can be derived in the lens programming
language {\em Boomerang}~\cite{boomerang}\bcp{All of them?  I'm not certain
  that's accurate.}. Each lens $\ell : S \Leftrightarrow T$
that Optician can synthesize specifies a bijection from the language of a
source regular expression $S$ to a target language of a target regular
expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations. Hence, Optician can be used as a
component in a system that synthesizes more complex lenses.

@@

For example, consider the following QRE for writing author
names:\bcp{Solomon: Someone noted on their copy of the intro from
  class yesterday that this example is backwards.  Did you go over the
  marked-up copies that I gave you?}
%
\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
   @Book {Lovelace,
   Author = "Ada Lovelace",
   Title = {Notes}
   }
\end{verbatim}
\noindent
and equivalent EndNote records like the following\bcp{bad punctuation: needs
a colon}
\begin{verbatim}
   %0 Book
   %F Lovelace
   %A Ada Lovelace
   %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:\bcp{bad punctuation}
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \collapse{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

The well-formedness constraints for QREs ensure that these four semantic
objects fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by \canonize(Q).

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function 
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.  
However,
verifying that a canonization
function $f$ is surjective and idempotent is in general undecidable.
Consequently, a programmer wishing to use \kw{normalize} must
discharge strong proof obligations, which is cumbersome in
practice.\footnote{In our implementation we allow a programmer to
use \kw{normalize} at their own risk without checking these side
conditions as an ``escape hatch'' for the case when other QRE
combinators are insufficient.\bcp{Do we use this in any of the
programs we've written?}\afm{Nope.}}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R))(w) &=& w \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f})(w) &=& f(w)\\
\canonize(Q_1 \; ; \; Q_2)(w) &=& \canonize(Q_2)(\canonize(Q_1)(w))\\
\canonize(Q_1 \cdot Q_2)(w_1 \cdot w_2) &=& \canonize(Q_1)(w_1) \cdot \canonize(Q_2)(w_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*)(\epsilon) &=& \epsilon \\
\canonize(Q^*)(w_1 \cdot w_2) &=& \canonize(Q)(w_1)\cdot\canonize(Q^*)(w_2) \\[2ex]
\multicolumn{3}{l}{\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})}\\[.5ex]
\multicolumn{3}{l}{\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1) \cdot \ldots \cdot
\canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)}
\end{array}
\]
\end{center}
\caption{QRE canonizers.  Here we assume that the input $w$ to the canonizers
  has been partitioned in the unique way guaranteed to exist by the lens unambiguity conditions.}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{l}
\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})\\[.5ex]
\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1) \cdot \ldots \cdot
\canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

First we give the formal definition of a {\em bijective quotient lens}; QRE
lenses are a specific class of bijective quotient lenses that map between data
that is specified by QREs. Given regular expressions $R, S$ and equivalence
relations $\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$,
we define a bijective quotient lens $q : R /{\equiv_R} \Leftrightarrow
S/{\equiv_S}$ to be a pair of functions $q.\get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.\lput : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r)) \equiv_R r$ and
$q.\get(q_1.\lput(s)) \equiv_R s$. Moreover, if $r \equiv_R r'$, then $q.\get(r) =
q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) = q.\lput(s')$.

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\equiv_R$
and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such that
$q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such that
the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$ and
$\mathcal{L}(S)/{\equiv_S}$ are mutual inverses.

These laws are similar to the laws for bijective lenses
\cref{bijectivelenslaws}\saz{again, cref uses a silly format}, except that the  equality restrictions in the
bijective lens laws are loosened to allow for equivalence relations. Also, the
condition that $r \equiv_R r'$ implies $q.\get(r) = q.\get(r')$ ensures that the
{\em get} function induced on the equivalence classes if $\equiv_c$ is
well-defined, and similarly for the {\em put} function.

@@

\subsection{Bijective String Lenses}
\label{subsec:bijective-lenses}
We define the set of \textit{bijective string lenses} to be the set of
bijections between regular languages constructed using the following
combinators.\bcp{bad punctuation}

$$\ell := \id(R) \sep \const(s, t) \sep  \swap(\ell_1, \ell_2)
\sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \;
\ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective string lenses, we now
introduce the class of QRE lenses. The language of QRE lenses is given by following
grammar:\bcp{bad punctuation}
$$ q := \lift(\ell) \sep  q_1 \cdot q_2 \sep \swap(q_1, q_2) \sep (q_1 \sep q_2)
\sep q^* \sep q_1 \; ; \; q_2 \sep \lquot(q, \ell) \sep \rquot(\ell, q),$$ where
$\ell$ ranges over bijective lenses.

@@

This rule essentially says that the composition $q_1 \; ; \; q_2$ is well
defined if and only if the intermediary QREs $Q_3$ and $Q_4$ define the same
equivalence relation on regular expressions that are equivalent modulo the
star-semiring axioms. (See~\cite[\S 4]{quotientlenses} for an example of what
goes wrong if this premise is dropped.\bcp{Didn't we used to have a discussion
  about how implementing this rule is hard because it involves checking language
  equivalence so we follow Boomerang's rule? I think this is important to
  mention because this was a significant restriction on the power of composition
  in Boomerang. It doesn't apply here (right?) because we intend to synthesize
  these lenses, not write them by hand, so we will never actually need to
  typecheck a functional composition expression.}\afm{Boomerang does check
  language equivalence, we restrict the check to star-semiring equivalence
  because we aren't complete when we allow arbitrary equivalence.}) The
condition $\mathcal{L}(W(Q_2)) = \mathcal{L}(W(Q_3))$ says that the intermediary
language is the same on both sides, while the condition $K(Q_2) \equiv^s K(Q_3)$
says that the kernel regular expressions are equivalent modulo the star-semiring
axioms. Finally, the condition $\canonize(Q_2) = \canonize(Q_3)$ says that $Q_2$
and $Q_3$ define the same equivalence relation on $\mathcal{L}(W(Q_2))=
\mathcal{L}(W(Q_3))$.

@@

By the induction hypothesis, there exist bijective lenses\bcp{every display
  for the rest of the section needs punctuation help}
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that
\begin{center}

\begin{tabular}{l l}
$\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1)$ & $\llbracket q_2 \rrbracket.\get = \llbracket \ell_2
\rrbracket \circ \canonize(Q_3)$ \\
$\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_2)$ & $\llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2
\rrbracket}^{-1} \circ \canonize(Q_4)$
\end{tabular}

\end{center}
%
Consequently,
\begin{align*}
\llbracket q_2 \rrbracket.\get \circ \llbracket q_1 \rrbracket.\get &=
(\llbracket \ell_2 \rrbracket \circ \canonize(Q_3)) \circ (\llbracket \ell_1
\rrbracket \circ \canonize(Q_1))\\
&= \llbracket \ell_2 \rrbracket \circ (\canonize(Q_3) \circ \llbracket \ell_1
\rrbracket) \circ \canonize(Q_1)\\
&= (\llbracket \ell_2 \rrbracket \circ \llbracket \ell_1 \rrbracket) \circ
\canonize(Q_1)\\
&= \llbracket \ell_1 \; ; \; \ell_2 \rrbracket \circ \canonize(Q_1)
\end{align*}
We are permitted to claim the third step from the second since
$\canonize(Q_3)$ is the identity function on $K(Q_2)$ which is
syntactically equal to $K(Q_3)$ by assumption. A similar argument shows that
$$\llbracket q_1 \rrbracket.\lput \circ \llbracket q_2 \rrbracket.\lput =
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket^{-1} \circ
\canonize(Q_4)$$

@@

Returning to the \bibtex{} to EndNote transformation of
Section~\ref{sec:examplesynth}, the QREs \cd{bibtex} and \cd{endnote}
describe a \bibtex{} record and an Endnote record respectively. We also had
the example pair \cd{(bib_example , end_example)} where\bcp{punctuation}

\begin{center}
\begin{minipage}{2.5in}
\begin{lstlisting}
bib_example =
"@Book {Lovelace,
 Author = \"Ada Lovelace\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{1.5in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A Ada Lovelace
 %F Lovelace"
\end{lstlisting}
\end{minipage}
\end{center}

There exists a bijective lens between the kernel of \cd{bibtex} and the kernel
of \cd{endnote} that maps the normalized form of \cd{bib_example} to the
normalized form of \cd{end_example}, so calling \textproc{SynthQRELens} on the
QREs \cd{bibtex} and \cd{endnote}, with the example set of
$\{(\text{\cd{bib\_example}},\text{\cd{end\_example}})\}$ will return a
satisfying lens. In this instance, the example set guides the algorithm to also
find the desired lens.

@@

$q_2 : Q_1 \Leftrightarrow Q_2$. Then

@@

By the induction hypothesis, there exist bijective lenses

@@

that

@@

Consequently,

@@

A similar argument shows that

@@

\Leftrightarrow K(Q_2)$ such that
that

@@

Consequentlty

@@

Similarly

@@

Then

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}(#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}
\newcommand{\bsep}{\ \ \sep{} \ \ }

@@

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats. A time-tested and appealing way to do this is
to use \emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction). The
main reason that lenses are appealing is because lens programming languages
guarantee that the \kw{put} and \kw{get} functions of a lens satisfy
certain \emph{lens laws} constraining how the two functions interact with each
other.

Recent work~\cite{optician} developed a tool called Optician to synthesize a
subset of the {\em bijective lenses} that can be derived in the lens
programming language {\em Boomerang}~\cite{boomerang}. Each lens $\ell : S
\Leftrightarrow T$ that Optician can synthesize specifies a bijection from the
language of a source regular expression $S$ to a target language of a target
regular expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations. Hence, Optician can be used as a
component in a system that synthesizes more complex lenses.

@@

For example, consider the following QRE for writing author
names:\bcp{Solomon: Someone noted on their copy of the intro from
  class yesterday that this example is backwards.  Did you go over the
  marked-up copies that I gave you?}\sam{I'm not sure what this comment means
  and also I forgot those notes in the office. I'll go through them more
  thoroughly tomorrow morning.}
%
\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
   @Book {Lovelace,
   Author = "Ada Lovelace",
   Title = {Notes}
   }
\end{verbatim}
\noindent
and equivalent EndNote records like the following:
\begin{verbatim}
   %0 Book
   %F Lovelace
   %A Ada Lovelace
   %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar,
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \collapse{R}{s} \bsep
\squash{Q_1}{Q_2}{f} \\ | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon
Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1 \sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

The well-formedness constraints for QREs ensure that these four semantic
objects fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by $\canonize(Q)$.

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function 
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.  
However,
verifying that a canonization
function $f$ is surjective and idempotent is in general undecidable.
Consequently, a programmer wishing to use \kw{normalize} must
discharge strong proof obligations, which is cumbersome in
practice.\footnote{In our implementation we allow a programmer to
use \kw{normalize} at their own risk without checking these side
conditions as an ``escape hatch'' for the case when other QRE
combinators are insufficient.}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R))(w) &=& w \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f})(w) &=& f(w)\\
\canonize(Q_1 \; ; \; Q_2)(w) &=& \canonize(Q_2)(\canonize(Q_1)(w))\\
\canonize(Q_1 \cdot Q_2)(w_1 \cdot w_2) &=& \canonize(Q_1)(w_1) \cdot \canonize(Q_2)(w_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*)(w_1 \cdot \ldots \cdot w_n) &=&
\begin{cases}
\epsilon & \text{if } n = 0\\
\canonize(Q)(w_1)\cdot \ldots \cdot \canonize(Q)(w_n) & \text{if } n > 0
\end{cases}
\\[2ex]
\multicolumn{3}{l}{\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot
s_1 \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})}\\[.5ex]
\multicolumn{3}{l}{\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1)
\cdot \ldots \cdot \canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)}
\end{array}
\]
\end{center}
\caption{QRE canonizers.  Here we assume that the input $w$ to the canonizers
  has been partitioned in the unique way guaranteed to exist by the lens unambiguity conditions.}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{l}
\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})\\[.5ex]
\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1) \cdot \ldots \cdot
\canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

Recall that given regular expressions $R, S$ and equivalence relations
$\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$, a
{\em bijective quotient lens} $q : R /{\equiv_R} \Leftrightarrow S/{\equiv_S}$
is a pair of functions $q.\get : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$
and $q.\lput : \mathcal{L}(S) \longrightarrow \mathcal{L}(R)$ such that for all
$r \in \mathcal{L}(R)$ and $s \in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r))
\equiv_R r$ and $q.\get(q_1.\lput(s)) \equiv_R s$. Moreover, if $r \equiv_R
r'$, then $q.\get(r) = q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) =
q.\lput(s')$. In words, a bijective quotient lens $q$ from $R$ to $S$ modulo
$\equiv_R$ and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such
that $q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such
that the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$
and $\mathcal{L}(S)/{\equiv_S}$ are mutual inverses. These laws are similar
bijective lens laws \cref{bijectivelenslaws}, except that the equality
restrictions in the bijective lens laws are loosened to allow for equivalence
relations. Also, the condition that $r \equiv_R r'$ implies $q.\get(r) =
q.\get(r')$ ensures that the {\em get} function induced on the equivalence
classes if $\equiv_c$ is well-defined, and similarly for the {\em put}
function.

@@

\subsection{Bijective String Lenses}
\label{subsec:bijective-lenses}
We define the set of \textit{bijective string lenses} to be the set of
bijections between regular languages constructed using the following
combinators,
$$\ell := \id(R) \sep \const(s, t) \sep  \swap(\ell_1, \ell_2)
\sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \;
\ell_1 \; ; \;  \ell_2$$
where $R$ ranges over regular expressions and $s$ ranges over character strings.

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective string lenses, we now
introduce the class of QRE lenses. The language of QRE lenses is given by
following grammar,
$$ q := \lift(\ell) \sep  q_1 \cdot q_2 \sep \swap(q_1, q_2) \sep (q_1 \sep q_2)
\sep q^* \sep q_1 \; ; \; q_2 \sep \lquot(q, \ell) \sep \rquot(\ell, q)$$
where $\ell$ ranges over bijective lenses.

@@

This rule essentially says that the composition $q_1 \; ; \; q_2$ is well
defined if and only if the intermediary QREs $Q_3$ and $Q_4$ define the same
equivalence relation on regular expressions that are equivalent modulo the
star-semiring axioms. (See~\cite[\S 4]{quotientlenses} for an example of what
goes wrong if this premise is dropped). The condition $\mathcal{L}(W(Q_2)) =
\mathcal{L}(W(Q_3))$ says that the intermediary language is the same on both
sides, while the condition $K(Q_2) \equiv^s K(Q_3)$ says that the kernel
regular expressions are equivalent modulo the star-semiring axioms. Finally,
the condition $\canonize(Q_2) = \canonize(Q_3)$ says that $Q_2$ and $Q_3$
define the same equivalence relation on $\mathcal{L}(W(Q_2))=
\mathcal{L}(W(Q_3))$.

Of course checking the condition $\canonize(Q_2) = \canonize(Q_3)$ is
undecidable in general; indeed the approach used to type quotient lenses in
Boomerang is to classify equivalences according to whether they are or are not
the equality relation. While this decision seemingly restricts the
power of composition in Boomerang significantly, the practice of writing
quotient lenses shows that this restriction is not overly restrictive. This is 
because most quotient lenses originate as lifted basic lenses, and
therefore have types whose equivalence relations are both equality, and further,
equality is preserved by many of the quotient lens combinators
~\cite{quotientlenses}. Foster et al also discuss a second possible approach to
typing quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice.

Thankfully, we do not face the issue of checking equivalence relation equality
in our work since our end goal is to synthesize lenses and not write them my
hand, so the programmer will never actually need to typecheck a functional
composition expression.

@@

By the induction hypothesis, there exist bijective lenses,
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that
\begin{center}:
\begin{tabular}{l l}
$\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1)$ & $\llbracket q_2 \rrbracket.\get = \llbracket \ell_2
\rrbracket \circ \canonize(Q_3)$ \\
$\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_2)$ & $\llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2
\rrbracket}^{-1} \circ \canonize(Q_4)$
\end{tabular}
\end{center}
%
Consequently:
\begin{align*}
\llbracket q_2 \rrbracket.\get \circ \llbracket q_1 \rrbracket.\get &=
(\llbracket \ell_2 \rrbracket \circ \canonize(Q_3)) \circ (\llbracket \ell_1
\rrbracket \circ \canonize(Q_1))\\
&= \llbracket \ell_2 \rrbracket \circ (\canonize(Q_3) \circ \llbracket \ell_1
\rrbracket) \circ \canonize(Q_1)\\
&= (\llbracket \ell_2 \rrbracket \circ \llbracket \ell_1 \rrbracket) \circ
\canonize(Q_1)\\
&= \llbracket \ell_1 \; ; \; \ell_2 \rrbracket \circ \canonize(Q_1)
\end{align*}
We are permitted to claim the third step from the second since
$\canonize(Q_3)$ is the identity function on $K(Q_2)$ which is
syntactically equal to $K(Q_3)$ by assumption. A similar argument shows that:
$$\llbracket q_1 \rrbracket.\lput \circ \llbracket q_2 \rrbracket.\lput =
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket^{-1} \circ
\canonize(Q_4)$$

@@

Returning to the \bibtex{} to EndNote transformation of
Section~\ref{sec:examplesynth}, the QREs \cd{bibtex} and \cd{endnote}
describe a \bibtex{} record and an Endnote record respectively. We also had
the example pair \cd{(bib_example , end_example)}, where:
\begin{center}
\begin{minipage}{2.5in}
\begin{lstlisting}
bib_example =
"@Book {Lovelace,
 Author = \"Ada Lovelace\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{1.5in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A Ada Lovelace
 %F Lovelace"
\end{lstlisting}
\end{minipage}
\end{center}
There exists a bijective lens between the kernel of \cd{bibtex} and the kernel
of \cd{endnote} that maps the normalized form of \cd{bib_example} to the
normalized form of \cd{end_example}, so calling \textproc{SynthQRELens} on the
QREs \cd{bibtex} and \cd{endnote}, with the example set of
$\{(\text{\cd{bib\_example}},\text{\cd{end\_example}})\}$ will return a
satisfying lens. In this instance, the example set guides the algorithm to also
find the desired lens.

@@

$q_2 : Q_1 \Leftrightarrow Q_2$. Then:

@@

By the induction hypothesis, there exist bijective lenses,

@@

that,

@@

Consequently:

@@

A similar argument shows that:

@@

\Leftrightarrow K(Q_2)$ such that:

@@

Consequentlty:

@@

Similarly:

@@

Then:$$
